---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.height = 4.5,
  fig.width = 4.5,
  dev = "ragg_png"
)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## moec.prioritizr: Multi-Objective Prioritization with the Epsilon-Constraint Approach

[![lifecycle](https://img.shields.io/badge/Lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/actions/workflow/status/AboozarM/moec.prioritizr/R-CMD-check-ubuntu.yaml?branch=master&label=Ubuntu)](https://github.com/AboozarM/moec.prioritizr/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/actions/workflow/status/AboozarM/moec.prioritizr/R-CMD-check-windows.yaml?branch=master&label=Windows)](https://github.com/AboozarM/moec.prioritizr/actions)
[![R-CMD-check-macOS](https://img.shields.io/github/actions/workflow/status/AboozarM/moec.prioritizr/R-CMD-check-macos.yaml?branch=master&label=macOS)](https://github.com/AboozarM/moec.prioritizr/actions)
[![Documentation](https://img.shields.io/github/actions/workflow/status/AboozarM/moec.prioritizr/documentation.yaml?branch=master&label=Documentation)](https://github.com/AboozarM/moec.prioritizr/actions)
[![Coverage Status](https://img.shields.io/codecov/c/github/AboozarM/moec.prioritizr?label=Coverage)](https://app.codecov.io/gh/AboozarM/moec.prioritizr/branch/master)

```{r "setup", include = FALSE}
# load developmental version of package
devtools::load_all()

# set default plot parameters
plot <- function(...) terra::plot(..., cex = 1.5)

# set up print method for prioritizr
print <- function(x, ...) {
  if (inherits(x, "MultiConservationProblem")) {
    prioritizr::knit_print.MultiConservationProblem(x)
  } else if (inherits(x, "ConservationProblem")) {
    prioritizr::knit_print.ConservationProblem(x)
  } else if (inherits(x, "OptimizationProblem")) {
    prioritizr::knit_print.OptimizationProblem(x)
  } else {
    base::print(x)
  }
}
```

```{r "knitr-config", include = FALSE}
# set parameters for figure sizes
h <- 4.5
w <- 6.5
ow <- "500"
is_github_readme <- isTRUE(knitr::pandoc_to()[[1]] %in% c("gfm", "markdown"))
knitr::opts_chunk$set(
  fig.height = h,
  fig.width = w,
  out.width = ow,
  fig.path = "man/figures/README-", fig.align = "center"
)
```

### Overview

Systematic conservation planning is a framework to identify priority areas for conservation management ([Giakoumi et al. 2025](https://doi.org/10.1016/j.tree.2024.12.002)). Briefly, priority areas are identified by formulating a mathematical optimization problem based on conservation objectives and requirements and then using optimization algorithms to generate solutions. To help account for multiple objectives, the _moec.prioritizr R_ package provides an implementation of the epsilon-constraint approach for multi-objective optimization ([Eichfelder 2008](https://doi.org/10.1007/978-3-540-79159-1)). The package is designed as a plugin for the [_prioritizr R_ package](https://prioritizr.net) -- a general purpose package for systematic conservation planning ([Hanson et al. 2025](https://doi.org/10.1111/cobi.14376)) -- and provide enhanced functionality. By using multi-objective optimization approaches, conservation scientists and practitioners can explore trade-offs and identify solutions that represent a desirable compromise among multiple objectives ([Neubert et al. 2025](https://doi.org/10.1016/j.tree.2025.09.007)).

### Installation

#### Package installation

The [latest developmental version of the _moec.prioritizr R_ package](https://github.com/AboozarM/moec.prioritizr) can be installed using the following _R_ code. Please note that it requires the [_sf_](https://CRAN.R-project.org/package=sf) and [_terra_](https://CRAN.R-project.org/package=terra) _R_ packages, which may require additional software to be installed. If you encounter problems installing these dependencies, please consult their installation instructions.

```{r "install", eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("AboozarM/moec.prioritizr")
```

### Usage

Here we provide a short example for using the _moec.prioritizr R_ package to generate conservation plans (hereafter, prioritizations) with the _prioritizr R_ package. Specifically, we will use an example dataset available through the _prioritizrdata R_ package. Additionally, we will use the _terra R_ package to perform raster calculations and the _ggplot2 R_ package for visualizations. To begin with, we will load the packages.

```{r "load-packages", message = FALSE, results = "hide"}
# to install packages for this example, please use:
# install.packages(c("prioritizr", "prioritizrdata", "terra", "ggplot2"))

# load packages
library(prioritizr)
library(prioritizrdata)
library(terra)
library(ggplot2)
```

We will use the Washington dataset in this example. To import the planning unit data, we will use the `get_wa_pu()` function. Although the _prioritizr R_ package can support many different types of planning unit data, here our planning units are represented as a single-layer raster (i.e., `terra::rast()` object). Each cell represents a different planning unit, and cell values denote land acquisition costs. Specifically, there are `r withr::with_options(list(scipen = 1000), as.character(terra::global(!is.na(get_wa_pu()), "sum", na.rm = TRUE)[[1]]))` planning units in total (i.e., cells with non-missing values).

```{r "planning_units"}
# import planning unit data
wa_pu <- get_wa_pu()

# preview data
print(wa_pu)

# plot data
plot(wa_pu, main = "Costs", axes = FALSE)
```

We will also import data on existing 0rotected areas. To achieve this, we will use the `get_wa_locked_in()` function to import spatially explicit protected area data as a single-layer raster, wherein cell values denote the locations of existing protected areas (values of 1 indicate existing protected areas).

```{r "locked-in"}
# import existing protected area data
wa_locked_in <- get_wa_locked_in()

# preview data
print(wa_locked_in)

# plot data
plot(wa_locked_in, main = "Existing protected areas", axes = FALSE)
```


We will now import data for native bird species. We will use the `get_wa_species()` function to import species distribution data as a multi-layer raster, wherein each layer corresponds to a different species and cell values denote the relative abundance of individuals (higher values mean greater abundance). To account for migratory patterns, the breeding and non-breeding distributions of species are represented as different layers. Additionally, we will use the `get_wa_attr()` function to import additional information for the species in tabular format (i.e., `tibble::tibble()` object), wherein each row corresponds to a different layer in the species distribution data. The `extinction_prob` column contains a probability of extinction and the `interest_score` column contains a measure of public interest for each species. Since the public interest scores contain numbers that range from zero to very high numbers, we will rescale these values to range between 0 and 1 to assist with interpretation.

```{r "species", fig.height = w * 1.2, fig.width = w * 1.5, out.width = "800"}
# import data
wa_species <- get_wa_species()
wa_attr <- get_wa_attr()

# rescale public interest scores,
# note we use log10 transformation due to very high values and add a
# value of 0.01 so that all species have a non-zero value
wa_attr$interest_score <- log10(wa_attr$interest_score + 1) + 0.01

# preview data
print(wa_species)
print(wa_attr)

# plot the first nine layers of species distribution data
plot(wa_species[[1:9]], nr = 3, axes = FALSE)
```

Next, we will import data for vulnerable carbon. Since habitat destruction through human activities can release sequestered carbon into the atmosphere, establishing protected areas in places with high vulnerable carbon can help reduce carbon emissions. We will use the `get_wa_carbon()` function to import spatially explicit data for vulnerable carbon as a single-layer raster, wherein cell values denote the total amount of carbon vulnerable to human activities.

```{r "carbon"}
# import data
wa_carbon <- get_wa_carbon()

# preview data
print(wa_carbon)

# plot data
plot(wa_carbon, main = "Carbon", axes = FALSE)
```

After importing the data, let's make sure that you have a solver installed on your computer. This is important so that you can use optimization algorithms to generate spatial prioritizations. If this is your first time using the _prioritizr R_ package, please install the HiGHS solver using the following _R_ code. Although the HiGHS solver is relatively fast and easy to install, please note that you'll need to install the [Gurobi software suite and the _gurobi_ _R_ package](https://www.gurobi.com/) for best performance (see the [Gurobi Installation Guide](https://prioritizr.net/articles/gurobi_installation_guide.html) for details).

```{r, eval = FALSE}
# if needed, install HiGHS solver
install.packages("highs", repos = "https://cran.rstudio.com/")
```

Now, let's generate a spatial prioritization to identify priority areas for protected areas establishment. In particular, we want to identify priority areas seek to promote representation of (i) threatened species (based on extinction probability, per `wa_attr$extinction_prob`), (ii) public interest species (based on public interest, per `wa_attr$interest_score`), and (iii) vulnerable carbon. When consider how well each species is represented, we will consider each species to be adequately represented if at least 40% of its distributed is covered by existing protected areas and priority areas (in other words, we will consider relative targets of 40%). Additionally, we will use locked in constraints so that existing protected areas are selected by the prioritization and ensure priority areas complement existing protected areas. Furthermore, we will set budgets to ensure that the total cost of the existing protected areas and priority areas does not exceed 3% of total costs across the study area. Since we have multiple conservation objectives that we want to consider, we will use the epsilon-constraint approach for multi-objective optimization. This particular approach is especially well-suited for generating a set of solutions to characterize a spectrum of trade-offs among multiple objective. For more information on building conservation planning problems (i.e., with `problem()` and `multi_problem()`), please refer to the [_prioritizr R_ package documentation](https://prioritizr.net).

```{r "problem", message = FALSE, results = "hide"}
# define a total conservation budget (3% of total cost)
budget <- 0.03 * terra::global(wa_pu, "sum", na.rm = TRUE)[[1]]

# create single-objective problem for threatened species
p1 <-
  problem(wa_pu, wa_species) %>%
  add_min_shortfall_objective(budget = budget) %>%
  add_relative_targets(0.8) %>%
  add_feature_weights(wa_attr$extinction_prob) %>%
  add_binary_decisions()

# create single-objective problem for public interest species
p2 <-
  problem(wa_pu, wa_species) %>%
  add_min_shortfall_objective(budget = budget) %>%
  add_relative_targets(0.8) %>%
  add_feature_weights(wa_attr$interest_score) %>%
  add_binary_decisions()

# create single-objective problem for vulnerable carbon,
# note that we only need to specify the constraints for one of the three
# problems because all constraints will be considered throughout the
# entire multi-objective optimization process
p3 <-
  problem(wa_pu, wa_carbon) %>%
  add_max_wtd_sum_objective(budget = budget) %>%
  add_locked_in_constraints(wa_locked_in) %>%
  add_binary_decisions()
```

```{r "multi_problem"}
# now create multi-objective problem with epsilon-constraint approach,
# and specify settings to generate a suite of different prioritizations
mp <-
  multi_problem(
    threatened_species = p1,
    public_interest_species = p2,
    vulnerable_carbon = p3
  ) %>%
  add_epconstraint_approach(n_per_problem = 3, verbose = TRUE) %>%
  add_default_solver(verbose = FALSE)

# print multi-objective problem
print(mp)
```

```{r "solve", message = FALSE, results = "hide"}
# solve problem to generate prioritizations,
# note that this will take a few minutes
ms <- solve(mp, run_checks = FALSE, remove_duplicates = TRUE)
```

After generating the prioritizations, we will now visualize the spatial distribution of priority areas.

```{r "maps", fig.height = w * 1.2, fig.width = w * 1.5, out.width = "800"}
# print the number of solutions generated
print(length(ms))

# convert prioritizations to multi-layer raster,
# and over-write values for existing protected areas to help with plotting
sols <- terra::rast(ms)
sols <- (sols * 2) - wa_locked_in

# plot the prioritizations as maps, wherein:
# cells with a value of 0 are note selected
# cells with a value of 1 are existing protected areas
# cells with a value of 2 are priority areas
plot(sols, nr = 3, axes = FALSE)
```

We can also calculate metrics to examine trade-offs between the different prioritizations.

```{r "metrics"}
# create table to store metrics for each prioritization,
# note we will use NA values as place-holder values to start with
metric_data <- data.frame(
  name = names(ms),
  threatened_species = NA,
  public_interest_species = NA,
  vulnerable_carbon = NA
)

# calculate metric to evaluate prioritizations according to their
# representation of threatened species (i.e., based on average
# percent of each species' target that is met, weighted by extinction
# probability)
metric_data$threatened_species <- vapply(
  ms,
  FUN.VALUE = numeric(1),
  function(x) {
    mean(
      eval_target_coverage_summary(p1, x)$relative_met *
      100 *
      wa_attr$extinction_prob
    )
  }
)

# calculate metric to evaluate prioritizations according to their
# representation of public interest species (i.e., based on average
# percent of each species' target that is met, weighted by public
# interest score)
metric_data$public_interest_species <- vapply(
  ms,
  FUN.VALUE = numeric(1),
  function(x) {
    mean(
      eval_target_coverage_summary(p1, x)$relative_met *
      100 *
      wa_attr$interest_score
    )
  }
)

# calculate metric to evaluate prioritizations according to their
# representation of vulnerable carbon (i.e., based on percentage of
# total vulnerable carbon covered by priority areas)
metric_data$vulnerable_carbon <- vapply(
  ms,
  FUN.VALUE = numeric(1),
  function(x) {
    mean(
      eval_feature_representation_summary(p1, x)$relative_held * 100
    )
  }
)

# print the metrics
print(metric_data)
```

```{r "trade-offs", fig.height = h, fig.width = w * 1.2, out.width = "800"}
# create plot to visualize trade-offs based on metrics
p <-
  ggplot(
    data = metric_data,
    aes(
      x = threatened_species,
      y = public_interest_species,
      color = vulnerable_carbon
    )
  ) +
  geom_point(size = 8) +
  scale_color_viridis_c() +
  labs(
    x = "Threatened species representation (%)",
    y = "Public interest species representation (%)",
    color = "Vulnerable carbon representation (%)",
  ) +
  theme(
    legend.position = c(0.99, 0.01),
    legend.justification = c(1, 0)
  )

# show plot
print(p)
```

### Citation

Please cite the _moec.prioritizr R_ package when using it in publications. To cite the package, please use:

```{r, echo = FALSE, results = "asis", comment = ""}
cat(paste0("> Mohammadi A and Hanson JO (2026). _moec.prioritizr: Multi-objective prioritization with the epsilon-constraint approach_. R package version 1.0.0, <https://AboozarM.github.io/moec.prioritizr/>."))
```

### Getting help

Please refer to the [package website](https://AboozarM.github.io/moec.prioritizr/) for more information. If you have any questions about using the package or suggestions for improving it, please [file an issue at the package's online code repository](https://github.com/AboozarM/moec.prioritizr/issues).
